Data Management \\ and LSST Special Programs

### 2017-07-31 ms.tex
updates after comments from ECB

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}

@@

It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible. Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream). Only images that qualify for Level 1 processing will contribute to the Alert Stream. The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template. So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine. However, the suitability of shorter and longer exposures is unknown. Here we consider the suitability -- or what we need to do in order to determine the suitability -- of shorter and longer exposures for the Level 1 and Alert Stream. These considerations are made independently of the science motivations of the Special Programs that proposed the different exposure times.

@@

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data. While it is expected that MOPS will be able to identify appropriate images and incorporate as much Special Programs data as possible, the capabilities are currently undefined, and we leave it for the MOPS pipeline builders to decide. We note that since MOPS does not run during the night, it does not have the same latency requirements, or the associated issues.

@@

$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1, the higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA. Is there a science need for 60 second latency on a consecutive time-series of short exposures, or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred? (I.e., the main goal of shorter exposures would be to include brighter stars, not necessarily to monitor variability on short timescales). Alternatively, are there options for short-term increases in parallel processing power at NCSA that we could invoke during times of high cadence observations? \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging. \end{enumerate}

@@

$\bullet$ Will near-real-time Level 3 difference-image processing be able to contribute to the same Alert Stream as Level 1? (E.g., transient source detections in nightly stacks of a deep drilling field.) \\
$\rightarrow$ \textcolor{blue}{Answered.} As described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream.

@@

$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used, but this will be needed for both Levels 1 and 3 differencing pipelines and products. \\
$\rightarrow$ \textcolor{red}{Unaddressed.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}

@@

$\bullet$ Will the following database element sizes be enough to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)? Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program, we need to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales. \\
{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}

@@

\noindent \textbf{Supernovae} \\
$\bullet$ Extend the SN sample to $z\sim1.2$ and providing more densely sampled light curves for cosmological analyses with the nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$. \\
%$\bullet$ \cite{CrottsWP} doesn't say much specific. \\
$\rightarrow$ The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$, which adds up to $<60$ minutes and could presumably be done with the standard 30s exposures. \cite{KesslerWP} \\
$\bullet$ Generating the best-possible individual SN light curves by building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will be necessary for SNe that appear in the template image (e.g., that last $>1000$ days). SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. \\
$\rightarrow$ The Level 2 DRP codes to create CoAdds should be adequate for this. \\
$\rightarrow$ \cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP. \\
$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but again this comes back to the short-exposure questions of whether we can do photometric calibration as well.

@@

Step 4. Level 3 Processing. \\ The federated pipeline runs automatically upon delivery of the Level 1 processed images. If a deep template image does not already exist (this would be a Level 3 product), a Level 2 stacking routine will be used to make them. This Level 2 stacking routine is also used to make the nightly deep CoAdd in each filter, and then a Level 1 or 2 routine will generate the difference images and run source detection (and real/bogus?). These sources will be added to Level 3 equivalents of the {\tt DIAObject} and {\tt DIASource} databases. 
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}

>>>>

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, NCSA about their processing capabilities, and DM about incorporating a functionality for the Alert Pipeline to go into a "skip intervening images" mode (or something similar), for $<30$ second cadence imaging. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used). \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? If so, its outputs might not be reliable because of train/test set mismatch; will the internal LSST routines for real/bogus training be available and adaptable to Level 3 users? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16a} Confirm that resources (packet distribution) for a Level 3 alert stream would be supported and with what latency, whether and how such packets would be identified as non-AP, and that they could be ingested by brokers in real time. \end{enumerate}

@@

It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible. Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST.). Only images that qualify for Level 1 processing will contribute to the Alert Stream. The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template. So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine. However, the suitability of shorter and longer exposures is unknown. Here we consider the suitability -- or what we need to do in order to determine the suitability -- of shorter and longer exposures for the Level 1 and Alert Stream. These considerations are made independently of the science motivations of the Special Programs that proposed the different exposure times.

@@

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data. Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline could be ingested by MOPS. We note that since MOPS does not run during the night, it does not have the same latency requirements, or the associated issues.

@@

$\bullet$ Special Programs that use short exposures (e.g., 5 seconds) pose a problem for the Level 1 Alert Pipeline because the higher data acquisition rate (e.g., $\sim$6$\times$ higher) would cause a processing back-up: the allocated processing resource is two parallel chains that take 60 seconds each, and so real-time processing only works for 30 second integrations. If there is a demonstrable science need for 60 second latency on each and every one of a consecutive series of short exposures, options for short-term increases in parallel processing power at NCSA will need to be considered (note that this may also be needed for crowded fields with >10k {\tt DIASources}). However, the main science goal of a short exposure Special Program may simply be to include brighter stars in the photometric catalogs, and not necessarily to monitor variability on short timescales. In this case, there is perhaps no science need to include each and every one of the exposures into the Alert Pipeline, and it might instead be useful enough to just include e.g., every 6$^{\rm th}$ image (which would need only the two parallel processing streams). There is currently no written functionality for this option: i.e., there is no way to send only every N$^{\rm th}$ image to the pipeline or, more flexibly, to skip over intervening images and only send an image once the processing pipeline is done with the last. \\
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, NCSA about their processing capabilities, and DM about incorporating a functionality for the Alert Pipeline to go into a "skip intervening images" mode (or something similar), for $<30$ second cadence imaging. \end{enumerate}

@@

$\bullet$ Will near-real-time Level 3 difference-image processing be able to contribute to the same Alert Stream as Level 1? (E.g., transient source detections in nightly stacks of a deep drilling field.) \\
$\rightarrow$ \textcolor{blue}{Answered.} As described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream (but a separate Level 3 alert stream would be possible).

@@

$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used, but this will be needed for both Levels 1 and 3 differencing pipelines and products. \\
$\rightarrow$ \textcolor{red}{Unaddressed.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used). \end{enumerate}

@@

$\bullet$ These are the database elements reserved for variability characterization: \\
{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
Will their sizes be adequate to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)? Keep in mind that we just need sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and that these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program. The concern here is just to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales. \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}

@@

\noindent \textbf{Supernovae} \\
$\bullet$ Extend the SN sample to $z\sim1.2$ and providing more densely sampled light curves for cosmological analyses with the nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$. \\
%$\bullet$ \cite{CrottsWP} doesn't say much specific. \\
$\rightarrow$ The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$, which adds up to $<60$ minutes and could presumably be done with the standard 30s exposures. \cite{KesslerWP} \\
$\bullet$ Generating the best-possible individual SN light curves by building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will be necessary for SNe that appear in the template image (e.g., that last $>1000$ days). SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. \\
$\rightarrow$ The Level 2 DRP codes to create CoAdds should be adequate for this. \\
$\rightarrow$ \cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP. \\
$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but again this comes back to the short-exposure questions of whether we can do photometric calibration as well.\\
$\bullet$ High-cadence observations to detect fast events, particularly young SNe and Gamma-Ray Bursts. \\

@@

Step 4. Level 3 Processing. \\ The federated pipeline runs automatically upon delivery of the Level 1 processed images. If a deep template image does not already exist (this would be a Level 3 product), a Level 2 stacking routine will be used to make them. This Level 2 stacking routine is also used to make the nightly deep CoAdd in each filter, and then a Level 1 or 2 routine will generate the difference images and run source detection (and real/bogus?). These sources will be added to Level 3 equivalents of the {\tt DIAObject} and {\tt DIASource} databases. The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream. 
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? If so, its outputs might not be reliable because of train/test set mismatch; will the internal LSST routines for real/bogus training be available and adaptable to Level 3 users? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16a} Confirm that resources (packet distribution) for a Level 3 alert stream would be supported and with what latency, whether and how such packets would be identified as non-AP, and that they could be ingested by brokers in real time. \end{enumerate}
